---
title: |
  STA 336: Statistical Machine Learning
subtitle: |
  Classwork 3
author: |
  Rento Saijo
date: 'April 4, 2026'
fontsize: 10pt
documentclass: extarticle
geometry: top=1.5in, bottom=1.5in, left=1.5in, right=1.5in
urlcolor: blue
linkcolor: blue
citecolor: blue
output:
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
    fig_width: 5
    fig_height: 3.5
    fig_caption: true

header-includes:
  - |
    \usepackage{xcolor}
    \usepackage{hyperref}
    \hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=blue}
    \usepackage{float}
    \usepackage{fvextra}
    \usepackage{xcolor}
    \usepackage{fancyhdr}
    \usepackage{lastpage}
    \usepackage{soul}
    \usepackage{etoolbox}
    \usepackage{microtype}
    \usepackage{tcolorbox}
    \usepackage{enumitem}
    \usepackage{fancyvrb}
    \usepackage{xspace}
    \allowdisplaybreaks
    \setcounter{tocdepth}{2}
    \setcounter{secnumdepth}{0}
    \floatplacement{figure}{H}
    \makeatletter
    \newcommand{\@subtitle}{}
    \newcommand{\subtitle}[1]{\gdef\@subtitle{#1}}
    \newcommand{\DocSubtitle}{\@subtitle}
    \renewcommand{\maketitle}{
      \thispagestyle{plain}
      \null
      \vfill
      \begin{center}
        {\Large \textsc{\@title} \par}\vskip 0.5em
        {\LARGE \bfseries \@subtitle \par}\vskip 0.75em
        {\large \@author \par}\vskip 0.5em
        {\normalsize \@date \par}
      \end{center}
      \vfill
      \tableofcontents
      \vspace{1em}
      \clearpage
    }
    \AfterEndEnvironment{Highlighting}{\setcounter{CodeLine}{\numexpr\value{CodeLine}+\FV@CodeLineNo\relax}}
    \makeatother
    \fancypagestyle{plain}{
      \fancyhf{}
      \renewcommand{\headrulewidth}{0pt}
      \renewcommand{\footrulewidth}{0pt}
    }
    \pagestyle{fancy}
    \fancyhf{}
    \fancyfoot[C]{\small\textsc{Page}~\thepage~\textsc{of}~\pageref{LastPage}}
    \fancyhead[L]{\textsc{Rento Saijo}}
    \fancyhead[R]{\textsc{\DocSubtitle}}
    \fancyhead[C]{\textsc{\rightmark}}
    \renewcommand{\headrulewidth}{0.5pt}
    \renewcommand{\footrulewidth}{0.5pt}
    \newcounter{CodeLine}
    \newcounter{cell}
    \AtBeginEnvironment{Highlighting}{\stepcounter{cell}}
    \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
      frame=single,
      rulecolor=\color{black},
      framesep=2mm,
      label=\footnotesize\textbf{Cell \thecell},
      labelposition=topline,
      commandchars=\\\{\},
      breaklines=true,
      breakanywhere=false,
      breaksymbol={},
      numbers=left,
      numbersep=3pt,
      firstnumber=\numexpr\value{CodeLine}+1\relax,
      fontsize=\small
    }
    \DefineVerbatimEnvironment{verbatim}{Verbatim}{
      breaklines=true,
      breakanywhere=true,
      numbers=left,
      numbersep=3pt,
      firstnumber=1,
      fontsize=\footnotesize
    }
    \DefineVerbatimEnvironment{snippet}{Verbatim}{
      breaklines=true,
      breakanywhere=true,
      fontsize=\footnotesize,
      frame=single,
      listparameters=\setlength{\topsep}{\baselineskip}\setlength{\partopsep}{0pt}
    }
    \newtcolorbox{problem}[1][]{
      title={\textsc{#1}}
    }
    \newenvironment{alphenum}{
      \begin{enumerate}[
        label=(\alph*),
        itemsep=3pt,
        parsep=0pt,
        topsep=6pt
      ]
    }{
      \end{enumerate}
    }
    \newenvironment{items}{
      \begin{itemize}[
        itemsep=3pt,
        parsep=0pt,
        topsep=6pt
      ]
    }{
      \end{itemize}
    }
    \renewcommand{\contentsname}{Table of Contents}
    \newcommand{\E}{\mathrm{E}}
    \newcommand{\Var}{\mathrm{Var}}
    \newcommand{\R}{\textsf{R}\xspace}
    \newcommand{\Pois}{\mathrm{Pois}}
    \newcommand{\SD}{\mathrm{SD}}
    \newcommand{\SE}{\mathrm{SE}}
---

```{r setup, include = FALSE}
# ----- Setup ----- #

# Set options.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
options(width = 125)
options(scipen = 999)

# Load libraries.
suppressMessages(library(magrittr))

# Set fonts.
sysfonts::font_add(
  family     = 'cmuserif',
  regular    = '/Users/rsai_91/Library/Fonts/cmu.serif-roman.ttf',
  bold       = '/Users/rsai_91/Library/Fonts/cmu.serif-bold.ttf',
  italic     = '/Users/rsai_91/Library/Fonts/cmu.serif-italic.ttf',
  bolditalic = '/Users/rsai_91/Library/Fonts/cmu.serif-bolditalic.ttf'
)
showtext::showtext_opts(dpi = 500)
showtext::showtext_auto(enable = TRUE)
knitr::opts_chunk$set(dev = 'CairoPDF')
par(family = 'cmuserif')
ggplot2::theme_set(ggplot2::theme_minimal(base_family = 'cmuserif', base_size = 10))

# ----- Helpers ----- #

# Print LaTeX table from data.frame.
table_latex <- function(
  data,
  col_names  = NULL,
  caption    = NULL,
  escape     = FALSE,
  align      = NULL,
  wrap_col   = NULL,
  wrap_width = NULL
) {
  if (is.null(align)) {
    align <- rep('c', ncol(data))
  }
  table_out <- knitr::kable(
    data,
    format    = 'latex',
    booktabs  = TRUE,
    escape    = escape,
    col.names = col_names,
    caption   = caption,
    linesep   = '',
    align     = align
  )
  table_out <- kableExtra::kable_styling(
    table_out,
    position      = 'center',
    latex_options = c('HOLD_position')
  )
  if (!is.null(wrap_col) && !is.null(wrap_width)) {
    table_out <- kableExtra::column_spec(table_out, wrap_col, width = wrap_width)
  }
  table_out
}

# Predict from regsubsets object.
predict.regsubsets <- function(object, newdata, id, ...) {
  form  <- stats::as.formula(object$call[[2]])
  mat   <- stats::model.matrix(form, newdata)
  coefi <- stats::coef(object, id = id)
  xvars <- names(coefi)
  mat[, xvars, drop = FALSE] %*% coefi
}

# Format numbers with commas.
comma_num <- function(x, digits = 0) {
  format(
    round(x, digits),
    big.mark   = ',',
    scientific = FALSE,
    trim       = TRUE,
    nsmall     = digits
  )
}

# Compute test MSE.
test_mse <- function(actual, predicted) {
  mean((actual - predicted)^2)
}

# Choose number of components from CV MSEP.
choose_ncomp <- function(fit) {
  msep <- drop(pls::MSEP(fit, estimate = 'CV')$val[1, 1, ])
  which.min(msep) - 1
}
```

# Disclosure

GPT-5.4 Codex was used to create the `YAML` portion and some `LaTeX` code to format the text/equations nicely. Page formatting code was also provided by Derin Gezgin. In the setup chunk, helper functions such as `table_latex()`, `predict.regsubsets()`, and `choose_ncomp()` were defined so the model-fitting output can be reported clearly and reproducibly. See the original `RMD` file [here](https://github.com/RentoSaijo/STA336/blob/main/Classwork3.Rmd) for more details.

\newpage

# Problem 1

\begin{problem}[Problem 1]
In this group work, we predict the number of applications received using the other variables in the \texttt{College} data set.
\end{problem}

\newpage

## Problem 1 (a)

\begin{problem}[Problem 1 (a)]
Split the data set into a training set and a test set.
\end{problem}

```{r, results = 'asis'}
# Load data.
college_df <- ISLR2::College

# Split data.
set.seed(24)
n         <- length(college_df$Apps)
train_idx <- sample(n, n / 2)
train_df  <- college_df[train_idx, ]
test_df   <- college_df[-train_idx, ]

# Summarize split.
split_tbl <- tibble::tibble(
  set          = c('Training', 'Test'),
  observations = c(nrow(train_df), nrow(test_df))
)

# Print split table.
split_tbl %>%
  table_latex(
    col_names = c('Set', 'Observations'),
    caption   = 'Training-test split for College.'
  )
```

Using `set.seed(24)`, I split the \(777\) observations in `College` into \(388\) training observations and \(389\) test observations. I use this same split for all four methods so that the resulting test errors are directly comparable.

\newpage

## Problem 1 (b)

\begin{problem}[Problem 1 (b)]
Fit a linear model using least squares on the training set, and report the test error obtained.
\end{problem}

```{r, results = 'asis'}
# Fit least squares model.
lm_fit  <- stats::lm(Apps ~ ., data = train_df)
lm_pred <- stats::predict(lm_fit, newdata = test_df)
lm_mse  <- test_mse(test_df$Apps, lm_pred)

# Summarize least squares result.
lm_tbl <- tibble::tibble(
  model    = 'Least Squares',
  test_mse = comma_num(lm_mse)
)

# Print least squares table.
lm_tbl %>%
  table_latex(
    col_names = c('Model', 'Test MSE'),
    caption   = 'Least squares test-set performance.'
  )
```

Fitting `Apps ~ .` on the training set gives a test MSE of \(\boxed{1{,}182{,}899}\).

\newpage

## Problem 1 (c)

\begin{problem}[Problem 1 (c)]
Fit a ridge regression model on the training set, with \(\lambda\) chosen by cross-validation. Report the test error obtained.
\end{problem}

```{r, results = 'asis'}
# Create model matrices.
x_train <- stats::model.matrix(Apps ~ ., data = train_df)[, -1]
y_train <- train_df$Apps
x_test  <- stats::model.matrix(Apps ~ ., data = test_df)[, -1]
y_test  <- test_df$Apps

# Fit cross-validated ridge model.
set.seed(24)
cv_ridge <- glmnet::cv.glmnet(
  x_train,
  y_train,
  alpha  = 0,
  lambda = seq(0, 25, by = 0.1),
  nfolds = 10
)

# Predict on test data.
ridge_pred <- stats::predict(cv_ridge, newx = x_test, s = 'lambda.min')
ridge_mse  <- test_mse(y_test, ridge_pred)

# Summarize ridge result.
ridge_tbl <- tibble::tibble(
  lambda   = comma_num(cv_ridge$lambda.min, 1),
  test_mse = comma_num(ridge_mse)
)

# Print ridge table.
ridge_tbl %>%
  table_latex(
    col_names = c('$\\lambda_{\\min}$', 'Test MSE'),
    caption   = 'Ridge regression test-set performance.'
  )
```

Using 10-fold cross-validation on the training set, ridge regression selects \(\lambda = 21.4\). The resulting test MSE is \(\boxed{1{,}255{,}479}\). On this split, ridge performs slightly worse than OLS.

\newpage

## Problem 1 (d)

\begin{problem}[Problem 1 (d)]
Fit a lasso model on the training set, with \(\lambda\) chosen by cross-validation. Report the test error obtained, along with the number of non-zero coefficient estimates.
\end{problem}

```{r, results = 'asis'}
# Fit cross-validated lasso model.
set.seed(24)
cv_lasso <- glmnet::cv.glmnet(
  x_train,
  y_train,
  alpha  = 1,
  lambda = seq(0, 25, by = 0.1),
  nfolds = 10
)

# Predict on test data.
lasso_pred <- stats::predict(cv_lasso, newx = x_test, s = 'lambda.min')
lasso_mse  <- test_mse(y_test, lasso_pred)

# Extract non-zero coefficients.
lasso_coef         <- as.matrix(glmnet::coef.glmnet(cv_lasso, s = 'lambda.min'))
lasso_nonzero_idx  <- which(lasso_coef[-1, 1] != 0) + 1
lasso_nonzero_tbl  <- tibble::tibble(
  term     = rownames(lasso_coef)[lasso_nonzero_idx],
  estimate = round(lasso_coef[lasso_nonzero_idx, 1], 6)
)
lasso_summary_tbl <- tibble::tibble(
  lambda              = comma_num(cv_lasso$lambda.min, 1),
  non_zero_predictors = nrow(lasso_nonzero_tbl),
  test_mse            = comma_num(lasso_mse)
)

# Print lasso summary.
lasso_summary_tbl %>%
  table_latex(
    col_names = c('$\\lambda_{\\min}$', 'Non-Zero Predictors', 'Test MSE'),
    caption   = 'Lasso test-set performance.'
  )
```

```{r, results = 'asis'}
# Print non-zero lasso coefficients.
lasso_nonzero_tbl %>%
  table_latex(
    col_names = c('Term', 'Estimate'),
    caption   = 'Non-zero lasso coefficients at $\\lambda_{\\min}$.'
  )
```

Lasso selects \(\lambda = 14.7\). Its test MSE is \(\boxed{1{,}234{,}483}\). At this value of \(\lambda\), there are \(13\) non-zero predictor coefficients (excluding the intercept), namely `PrivateYes`, `Accept`, `Top10perc`, `Top25perc`, `F.Undergrad`, `P.Undergrad`, `Outstate`, `Room.Board`, `PhD`, `Terminal`, `perc.alumni`, `Expend`, and `Grad.Rate`.

\newpage

## Problem 1 (e)

\begin{problem}[Problem 1 (e)]
Fit a PCR model on the training set, with \(M\) chosen by cross-validation. Report the test error obtained, along with the value of \(M\) selected by cross-validation.
\end{problem}

```{r, results = 'asis'}
# Fit PCR with cross-validation.
set.seed(24)
pcr_fit <- pls::pcr(Apps ~ ., data = train_df, scale = TRUE, validation = 'CV')
pcr_m   <- choose_ncomp(pcr_fit)

# Predict on test data.
pcr_pred <- stats::predict(pcr_fit, newdata = test_df, ncomp = pcr_m)
pcr_mse  <- test_mse(y_test, pcr_pred)

# Summarize PCR result.
pcr_tbl <- tibble::tibble(
  selected_m = pcr_m,
  test_mse   = comma_num(pcr_mse)
)

# Print PCR table.
pcr_tbl %>%
  table_latex(
    col_names = c('Selected $M$', 'Test MSE'),
    caption   = 'PCR test-set performance.'
  )
```

PCR chooses \(\boxed{M = 17}\) principal components, and the resulting test MSE is \(\boxed{1{,}182{,}899}\). This exactly matches the least squares test error on this split. That agreement makes sense here because `College` has \(17\) predictors, so using all \(17\) principal components reproduces the full least squares fit.

\newpage

## Problem 1 (f)

\begin{problem}[Problem 1 (f)]
Comment on the results obtained. How accurately can we predict the number of college applications received? Is there much difference among the test errors resulting from these four approaches?
\end{problem}

```{r, results = 'asis'}
# Create comparison table.
comparison_tbl <- tibble::tibble(
  model    = c('Least Squares', 'Ridge', 'Lasso', 'PCR'),
  tuning   = c(
    'None',
    paste0('$\\lambda = ', comma_num(cv_ridge$lambda.min, 1), '$'),
    paste0('$\\lambda = ', comma_num(cv_lasso$lambda.min, 1), '$'),
    paste0('$M = ', pcr_m, '$')
  ),
  test_mse = c(
    comma_num(lm_mse),
    comma_num(ridge_mse),
    comma_num(lasso_mse),
    comma_num(pcr_mse)
  )
)

# Print comparison table.
comparison_tbl %>%
  table_latex(
    col_names = c('Model', 'Tuning', 'Test MSE'),
    caption   = 'Comparison of the four required methods.',
    escape    = FALSE
  )
```

Among the four required approaches, least squares and PCR tie for the best test performance, both with test MSE \(1{,}182{,}899\). Lasso is slightly worse at \(1{,}234{,}483\), and ridge is worst at \(1{,}255{,}479\). The gap between the best and worst methods is \(72{,}580\) in test MSE, so there is some difference, but not a dramatic one. Overall, the models predict `Apps` reasonably well, but the shrinkage methods do not improve over least squares here, and PCR also does not help because cross-validation selects all \(17\) components.

\newpage

## Problem 1 (g)

\begin{problem}[Problem 1 (g)]
If you have more time, what else can you try to build a model for predicting the number of applications received? Can it improve the models you have built?
\end{problem}

```{r, results = 'asis'}
# Fit best subset models.
best_subset_fit <- leaps::regsubsets(Apps ~ ., data = train_df, nvmax = 17)
best_subset_mse <- sapply(1:17, function(i) {
  pred <- predict.regsubsets(best_subset_fit, test_df, id = i)
  test_mse(y_test, pred)
})

# Extract best subset result.
best_subset_size <- which.min(best_subset_mse)
best_subset_tbl  <- tibble::tibble(
  model_size = best_subset_size,
  predictors = paste(names(stats::coef(best_subset_fit, id = best_subset_size))[-1], collapse = ', '),
  test_mse   = comma_num(min(best_subset_mse))
)

# Print best subset extension.
best_subset_tbl %>%
  table_latex(
    col_names  = c('Model Size', 'Predictors', 'Test MSE'),
    caption    = 'Best subset extension on the same split.',
    align      = c('c', 'l', 'c'),
    wrap_col   = 2,
    wrap_width = '8cm'
  )
```

One natural extension, since we used it in class, is best subset selection with `leaps::regsubsets()`. On this same split, the best subset approach actually improves on the four required models: the lowest test error comes from the 1-variable model using only `Accept`, and its test MSE is \(\boxed{1{,}093{,}185}\). This is \(89{,}714\) lower than the best test MSE among the required four methods, so in this case, yes, trying an additional model-selection approach does improve predictive performance.
